Abundant Parent-of-origin Effect eQTL: The Framingham Heart Study

Version 1

Abstract

Parent-of-origin effect (POE) is a phenomenon whereby an allele’s effect on a phenotype depends both on its allelic identity and parent from whom the allele is inherited, as exemplified by the polar overdominance in the ovine callypyge locus and the human obesity DLK1 locus. Systematic studies of POE of expression quantitative trait loci (eQTL) are lacking. In this study we use trios among participants in the Framingham Heart Study to examine to what extend POE exists for gene expression of whole blood using whole genome sequencing and RNA sequencing. For each gene and the SNPs in cis, we performed eQTL analysis using genotype, paternal, maternal, and joint models, where the genotype model enforces the identical effect sizes on paternal and maternal alleles, and the joint model allows them to have different effect sizes. We compared models using Bayes factors to identify paternal, maternal, and opposing eQTL, where paternal and maternal effects have opposite directions. The resultant variants are collectively called POE eQTL. The highlights of our study include: 1) There are more than 2, 000 genes harbor POE eQTL and majority POE eQTL are not in the vicinity of known imprinted genes; 2) Among 180 genes harboring opposing eQTL, 99 harbor exclusively opposing eQTL, and 58 of the 99 are phosphoprotein coding genes, reflecting significant enrichment; 3) Paternal eQTL are enriched with GWAS hits, and genes harboring paternal eQTL are enriched with drug targets. Our study demonstrates the abundance of POE in gene expression, illustrates the complexity of gene expression regulation, and provides a resource that is complementary to existing resources such as GTEx. We revisited two previous POE findings in light of our POE results. A SNP residing in KCNQ1 that is maternally associated with diabetes is a maternal eQTL of CDKN1C, not KCNQ1. A SNP residing in DLK1 that showed paternal polar overdominance for human obesity is a maternal eQTL of MEG3, offering an explanation for the baseline risk of homozygous samples through association between MEG3 expression and obesity. Finally, we advised caution on conducting Mendelian randomization using gene expression as the exposure.

Background

Parent-of-origin effect (POE) is a phenomenon whereby an allele’s effect on a phenotype not only depends on its allelic identity, but also on the parent from whom the allele is inherited (DeChiara et al., 1991; Cockett et al., 1996). POE can be driven by factors such as sex bias in transmission of genetic variants (Tomé et al., 2011) and maternal genetic effects that influences the environment in which the offspring develops (Hager et al., 2008). The main driver, however, is genomic imprinting, which is a phenomenon where one of the two alleles at a locus is functionally silenced by methylation. Importantly, which parental copy of the allele being silenced is highly consistent for that gene across samples. Imprinting appears to play an important role in embryonic and placental development and social behavior (Reik and Walter, 2001; Garfield et al., 2011). Genes underpinning syndromic disorders such as Prader-Willi, Beckwtih-Weidemann, and Angelman show POE (Lawson et al., 2013). The same is also true for common disease phenotypes, such as breast cancer, diabetes, and cardiovascular disease (Kong et al., 2009; Hanson et al., 2013; Mozaffari et al., 2019).

In genetic association studies, particularly genome-wide association studies (GWAS), the absence of POE is usually assumed implicitly. This assumption is largely out of convenience, as in most GWAS, the parent-of-origin of an allele is not directly observable, and not inferable due to lack of first and second degree relatives in the sample. Exceptions include an Icelandic cohort, with their extensively documented pedigrees and the abundance of closely related samples (Kong et al., 2009), a Hutterite pedigree (Mozaffari et al., 2019), the Framingham Heart Study with three-generations of participants (Kannel et al., 1979), and the biobank datasets, for which parent-of-origin can be inferred using distant relatives (Hofmeister et al., 2022). Another contributing factor to the implicit assumption may be the presumed scarcity of documented POE in humans. Although POE is believed to be associated with a wide range of complex traits and diseases (Lawson et al., 2013), significant associations reported in GWAS setting are of limited quantity. For example, in a Hutterite study, Mozaffari et al. (2019) examined 21 phenotypes and produced 18 significant associations with POE, fewer than one association per phenotype.

Mouse studies demonstrated that, despite a limited number of imprinted genes, most phenotypes display POE, and loci that show POE were not enriched for known imprinted genes. Using reciprocal F1 crosses, Mott et al. (2014) showed that non-imprinted genes can generate POE by interactions with imprinted loci. Focusing on metabolic traits and using a mouse population at different levels of intercrossing, Macias-Velasco et al. (2022) identified a network comprised of three imprinted and six non-imprinted genes that show POE. In humans, however, loci with POE in relation to phenotypes are either near imprinted regions by design (Kong et al., 2009; Hanson et al., 2013), or near regions with characteristic of imprinting from a genome-wide scan (Mozaffari et al., 2019). To what extend POE contributes to phenotypic variation in humans and how widespread the loci with POE in the human genome remains elusive.

Using whole genome sequencing and RNA sequencing data from the Framingham Heart Study, and taking advantage of its relative abundance of trios, we set to investigate to what extent and degree gene expressions is affected by parent-of-origin. There exist multiple studies on POE on gene expression (Zhabotynsky et al., 2019; Jadhav et al., 2019; Deng et al., 2020) or POE of methylation and gene expression jointly (Zink et al., 2018), but these studies are either underpowered due to small sizes or focus on analysis of allele specific gene expression instead of eQTL. A systematic study of POE eQTL is lacking and our study aims to fill this void.

Results

Overview of data processing

We identified 1477 trios with whole genome sequencing (WGS) data and children within the trios having RNAseq data. After routine QC of genotype data, for each trio, we first phased the child based on rules of Mendelian inheritance, which resolved all non triple heterozygous genotypes. The genotype data has negligible Mendelian error rate, and see details in Methods on how to handle them.

We then marked the remaining triple heterozygous genotypes as missing, which produced phased paternal and maternal haplotypes for each child in the trio. Next we imputed missing alleles into their paternal and maternal haplotype background, and the parental haplotype with a higher imputed allele dosage was assigned the reference allele, the other haplotype was assigned the alternative allele. Simulation studies showed this mask and impute phasing to be highly accurate (Supplementary Table S1). Thus at each SNP, we have a genotype vector, a paternal vector, and a maternal vector.

To process RNAseq data, we started from Transcript Per Million (TPM) values, and selected 16, 824 genes that have < 5% of 0 TMP values. We then corrected for the GC content bias using an approach that is similar to the one used by EDAseq (Risso et al., 2011). The corrected gene expression values were then quantile normalized.

Bayes factors as evidence for association

For genetic association we used a linear mixed model, fitted by a novel and efficient method that is designed for analyzing multiomics datasets (Guan and Levy, 2024a). The genetic relatedness matrix (twice of kinship matrix) used for the linear mixed model was estimated by Kindred (Guan and Levy, 2024c). The covariates included age, sex, body mass index, and white blood cell composition. This study focuses on genetic association between gene expression and genetic variants in cis, defined as within 1Mb from transcription start site. We elected to use Bayes factors (Kass and Raftery, 1995) instead of p-values as the evidence of association for the following considerations: 1) Bayes factor has been successfully applied in high profile studies as evidence for genetic association (Wellcome Trust Case Control Consortium, 2007). 2) Bayes factor and p-values have a nice monotone relationship for a fixed allele frequency (Zhou and Guan, 2018; Guan and Levy, 2024b). If an eQTL is significant with Bayes factor, it is likely also significant with p-value, but not vice versa (due to the next point). 3) Bayes factors panelize genetic variants with small allele frequencies to reduce false positives (Guan and Stephens, 2011; Zhou and Guan, 2018). In other words, variants of small allele frequencies may have significant p-values, but unsignificant Bayes factors. This is useful as paternal or maternal allele frequencies can be low by chance even the genotype minor allele frequencies are chosen to be ≥ 0.01. 4) Informally, Bayes factor accounts for multiple testing through prior odds in a manner similar to Bonferroni correction: a smaller prior odds requires larger Bayes factor to maintain the same level of posterior odds for association. 5) Parent-of-origin effects require comparing multiple non-nested modles (e.g., maternal vs. paternal effects). Bayes factors provide a coherent framework for such comparisons, while frequentist approaches lack an equivalent (Stephens and Balding, 2009).

For each gene and its cis-SNP, we computed four Bayes factors: genotype Bayes factor (BFg), paternal Bayes factor (BF1), maternal Bayes factor (BF0), and joint Bayes factors (BFj). When computing joint Bayes factor, we also computed paternal and maternal effect sizes and attached a p-value to test whether they differ significantly. Examples of test statistics are provided in Table 1. BFg, BF0 and BF1 are one degree of freedom tests, but BFj is a two degree of freedom test. Thus to achieve BFj > BFg, the paternal and maternal effect sizes have to differ significantly to compensate for the extra degree of freedom. Examples in Table 1 are chosen because their paternal and maternal effects differ significantly, and we have BFj >> BFg. Another useful observation is that BF1 is correlated with |β1|, and BF0 is correlated with |β0|. But BF1 and BF0 are oblivious to the signs of β1 and β0, and signs are informative and interesting in our context.

We used log10 BF = 4 as the significance threshold. If the prior odds for a cis-eQTL is 1 out of 1000, this threshold gives the posterior probability of association (PPA) of 0.91. If the prior odds for a cis-eQTL is 1 out of 100, this threshold gives the PPA of 0.99. 1 – PPA is the Bayesian counterpart of the local false discovery rate (c.f. Soloff et al., 2024).

Overview from sentinel eQTL of joint analysis

Table 1 contains 10 genes and test statistics of their sentinel eQTL. These eQTL are chosen because they have most significant column P (see caption for details). A sentinel eQTL for a gene is defined as the eQTL with largest BFj for that gene. A gene with significant eQTL is called eGene. These sentinel eQTL are either BF1 >> BF0 ≈ 1 such as NDN, or BF0 >> BF1 ≈ 1 such as MEG3, or BFj >> BFg ≈ 1 such as NECAB3. In Table 1 the sentinel eQTL were ordered according to column P, which measures the significance of differences between paternal and maternal effects. Reassuringly, six out of top ten sentinel eQTL with most significant P are from bona fide imprinted genes (colored in blue) according to http://geneimprint.com.

Figure 1 plots paternal effect vs maternal effect for all sentinel eQTL whose log10 BFj > 3. The color of dots corresponding to degree of difference between paternal and maternal effects. The marked sentinel eQTL include known imprinted genes such as maternally expressed ZNF331 and MEG3, and paternally expressed FAM50B, GNAS, SGCE, NDN, SNURF, SNRPN, and PEG10. Imprinted genes are located along the x and y-axes. At least two genes appear to be novel imprinted genes: PPIEL and CCR9. Gene ZNF890P provides an example of paternal and maternal effects being in the same direction, but their sizes differ significantly. Finally, NECAB3 and LSM7 are two examples of paternal and maternal effects that are in opposite directions.

Comparison of paternal effects and maternal effects of sentinel eQTL in joint analysis.

Each point is the most prominent eQTL with log10 BFj > 3 of gene. The paternal effect is on x-axis and maternal effect y-axis. The coloring reflects significance of differentials between paternal effects and maternal effects, with red more significant than blue.

This approach based on the sentinel eQTL suggests that 1) the combined effect of paternal and maternal allele can be across all 360 degrees; 2) when paternal and maternal effects are in opposite directions, both effects sizes are modest compared to when they are in the same direction; and 3) there are quite a few sentinel eQTL whose paternal and maternal effects show opposite directions.

NECAB3 : opposite paternal and maternal effects

The existence of eGenes whose eQTL show opposite paternal and maternal effects is intriguing. We focused on the example of NECAB3 to examine further details. Figure 2 left panel shows that all eQTL of NECAB3 have opposite paternal and maternal effects. Consequently, the joint Bayes factors BFj’s are much larger than their corresponding genotype Bayes factors BFg’s (right bottom).

An example of gene that whose paternal and maternal eQTL have opposite effect sizes.

On the left is paternal vs maternal effect sizes normalized by their corresponding standard deviation. The effect sizes were estimated by joint analysis. The non-significant eQTL are colored in gray and significant eQTL colored in black. An eQTL marked in blue was chosen to show boxplots with gene expression (phenotype) of NECAB3. The three panels of boxplots are for genotypes, paternal alleles, and maternal alleles. For this SNP rs4911348, the genotypes has no association with phenotype, but both paternal alleles and maternal alleles are associated with the phenotype, and the paternal and maternal effects are in opposite direction. Right panel bottom show Manhattan plot of cis-eQTL of the gene NECAB3. The genotype Bayes factor were colored in gray and joint test Bayes factor were colored in black. SNP rs4911348 was highlighted in blue.

We selected as an example SNP rs4911348, which is different from the sentinel eQTL shown in Figure 1, and looked into details of its association with the gene expression of NECAB3 (phenotype). Three panels of boxplots (right top) show that genotype has no association with the phenotype, and both paternal alleles and maternal alleles are associated with the phenotype, but in opposite directions. Opposite paternal and maternal effects were also observed in Hutterite POE study, where ten associations of opposite effects across nine traits were reported (Mozaffari et al., 2019).

POE eQTL are abundant

Next we looked beyond the sentinel eQTL and examined all significant eQTL. We noted that there are 15, 893 eQTL from 14, 733 SNPs and 1, 824 eGenes that are insignificant for the SNP test, but significant for the paternal, or maternal, or joint test. That is, log10 BFg < 4, but either log10 BF1 > 4, or log10 BF0 > 4, or log10 BFj > 4. In other words, these eQTL cannot be detected without analyzing POE, which implies that POE can contribute to recover missing heritability.

With reference Figure 1, we proposed criteria to define subsets of eQTL. The first subset eQTL locate along the x-axis, mimicking sentinel eQTL of NDN and FAM50B; the second subset locate along the y-axis, mimicking sentinel eQTL of ZNF331 and MEG3; the third subset locate along the secondary diagonal, mimicking sentinel eQTL of NECAB3 and LSM7; and the fourth subset follows (blues dots) along the diagonal line. Note Figure 1 only involves BFj, our criteria to define gene sets also takes into account of BF1, BF0, and BFg (details in Methods).

The first set of eQTL have significant paternal effects but insignificant maternal effects. We identified 15, 576 such paternal eQTL (set SP) from 14, 372 SNPs associated with 1, 188 eGenes. The second set of eQTL have significant maternal effects but insignificant paternal effects. We identified 14, 783 such maternal eQTL (set SM) from 13, 293 SNPs associated with 1, 209 eGenes. We refer to these paternal and maternal eQTL as imprinting eQTL. The third set of eQTL have opposite paternal and maternal effects, such that joint Bayes factors are much larger than genotype Bayes factor BFj >> BFg. We identified 688 such opposing eQTL (set SO) from 485 SNPs that are associated with 180 eGenes. Imprinting eQTL and opposing eQTL are referred to as POE eQTL.

The fourth set of eQTL require that both paternal and maternal alleles alone are associated with an expression phenotype, and the two effects are in the same direction. For these eQTL BFg > BFj because similar effect sizes favor one degree of freedom test. We identified 884, 119 such genotype eQTL from 577, 701 SNPs and 4940 eGenes. The percent of eGenes with genotype eQTL is on par, but smaller than GTEx study (Consortium et al., 2020), presumably because log10 BFg > 4 is a more stringent threshold for eQTL analysis.

Paternal eQTL are enriched with GWAS hits

Using ANNOVAR (Wang et al., 2010), we annotated all four sets of eQTL (Table 2). We made the following observations: 1) The median distance to transcription start site (TSS) for opposing eQTL is 673.2Kb, much larger than other types of eQTL, and majority of opposing eQTL SO are located in introns of nearby genes. 2) The percent of GWAS hits for SNPs in SO is 0.022, significantly lower than that of SNPs in genotype eQTL SG’s 0.054 (test for proportion P = 0.001). This is reasonable as GWAS mainly use genotype test and are agnostic to opposing eQTL. 3) The precent of GWAS hits for SNPs in paternal eQTL SP is higher than that of SNPs in SG (test for proportion P = 9 × 10−11), and the percent of GWAS hits among SNPs in maternal eQTL SM has no significant difference to that of SNPs in SG (test for proportion P = 0.051).

Number of eQTL per SNP (r) measures degree of pleiotropy of a set of SNPs. SNPs in SP and SNPs in SM have similar pleiotropy r ≈ 1.1, SNPs in SG have the largest r = 1.53, and SNPs in SO display r = 1.41. In other words, SNPs are more exclusive for imprinting eQTL, less exclusive for genotype eQTL, and somewhere in between for opposing eQTL. Table 2 (bottom) compares how the degree of pleiotropy affects their annotation for SNPs in SG, where set Pk contains SNPs in SG that are eQTL of at least k eGenes. The most interesting observation is that the percent of GWAS hits increases with rising SNP pleiotropy. This feature is only observed in genotype eQTL.

POE eGenes are abundant

There are 1, 188 eGenes harboring paternal eQTL and 1, 209 eGenes harboring maternal eQTL. Their union contains 2, 139 eGenes harbor imprinting eQTL, and their intersection contains 258 eGenes harboring both paternal and maternal eQTL. There are 4, 940 eGenes harboring genotype eQTL, and 1, 867 eGenes harboring both genotype eQTL and imprinting eQTL. Supplementary Figure S1 showed an example of an eGene harboring both maternal and paternal eQTL: GZMH, and an example of an eGene harboring both genotype eQTL and imprinting eQTL: DSE. The most prominent feature is that paternal eQTL and maternal eQTL cluster in separate genomic regions, and genotype eQTL and imprinting eQTL also cluster in separate genomic region.

eGenes that harbor exclusively imprinting eQTL

There are 129 eGenes (set G1) that harbor exclusively paternal eQTL, and 139 eGenes (set G0) that harbor exclusively maternal eQTL. Naturally, G1 contain many bona fide imprinted genes with paternal expression such as FAM50B, NDN, SGCE, SNRPN, SNURF and PEG10. Remaining genes in G1 are candidates for imprinted genes with paternal expression, for example CCR9 and NCOA2. G0 contain bona fide imprinted genes with maternal expression such as GRB10, ZNF331 and MEG3. The remaining in G0 are candidates for imprinted genes with maternal expression, for example COA8 and ZNF888.

The Supplementary Material provides a full list of the candidate imprinted genes from the eQTL analysis. Note however, that not all imprinted genes have eQTL, therefore this list is incomplete. Moreover, a gene on the list is considered behaving like an imprinted gene, but may not be a bona fide imprinted gene. For example, a gene that has no cis-eQTL on its own, may “acquires” a cis-eQTL through gene-gene interaction with a putative imprinted gene such that the cis-eQTL is in fact its trans-eQTL.

We used G1 and G0 as proxies for paternal and maternal imprinted genes, and used cytoBands these genes occupies as proxies for imprinted regions to examine to what extend paternal and maternal eQTL locate in the same cytoBands as these proxy imprinted genes. There are 811 autosome cytoBands in humans, 765 of them larger than 1 Mb, and the median size is 3.2 Mb (Cheung et al., 2001). The 129 proxy paternal imprinted genes occupy 89 cytoBands (set ), while 15, 576 paternal eQTL occupy 445 cytoBands (set ), among them 9, 947 (or 64%) were not in the cytoBands set . On the other hand, the 139 proxy maternal imprinted genes occupy 96 cytoBands (set ), and 14, 783 maternal eQTL occupy 465 cytoBands (set ), among them 7, 987 (or 54%) were not in the cytoBand set . Therefore, a majority of imprinting eQTL are not in the vicinity of the proxy imprinted regions. Interestingly, and share only 14 cytoBands, 8% of their union of 171 cytoBands. As a comparison, and share 349 cytoBands, 62% of their union of 561 cytoBands.

eGenes that harbor exclusively opposing eQTL

There are 99 eGenes (set G2) that harbor only opposing eQTL (Supplementary). Among them 58 encode phosphoproteins (Supplementary), a significant enrichment (FDR = 5.5 × 10−4) according to DAVID, a web server for functional enrichment analysis (Sherman et al., 2022). A phosphoprotein is a protein that is posttranslationally modified by the attachment of either a single phosphate group, or a complex molecule such as 5’-phospho-DNA, through a phosphate group. Because phosphorylation often serves as an on/off switch, targeting phosphoproteins or the enzymes regulating them can restore normal cell signaling in diseases. In addition, drugs can modulate the functions of these proteins, either inhibiting or enhancing their activities. This makes phosphoprotein attractive drug targets. Indeed, according to a therapeutic target database (Zhou et al., 2024), among 58 phosphoproteins genes that harbor exclusively opposing eQTL, there are three successful drug targets ITPR1, ITPR2, and REL, and three additional clinical trial targets CD46, NOTCH3, and XPO1.

Other notable phosphoproteins genes include CDKN2A, PHIP, and LEP, among others. Where CDKN2A produces two major proteins p16(INK4), which is a cyclin-depnednet kinase inhibitor, and p14(ARF), which binds the p53-stabilizing protein MDM2 (Serrano et al., 2000); PHIP is associated with Chung-Jansen syndrome, featuring behavioral problems, intellectual disability, obesity, and dysmorphia (Webster et al., 2016; Jansen et al., 2018); and LEP encodes leptin, a protein that plays a critical role in the regulation of body weight. Leptin is secreted by white adipocytes, and it binds to the leptin receptor in the brain, which in turn inhibits appetite and promotes energy expenditure (Maffei et al., 1995; Weigle et al., 1997).

Paternal eGenes are enriched with drug targets

We used a therapeutic target database (Zhou et al., 2024) to examine the enrichment with drug targets of the four sets of eGenes, grouped by their eQTL. Table 3 contains the enrichment of the four gene sets in either successful targets, clinical trial targets, or combined targets. The significant enrichment that (following Bonferroni correction of 12 tests) is highlighted. Both paternal eGenes and genotype eGenes are enriched in clinical trial targets and combined targets, but paternal eGenes are significantly more enriched than genotype eGenes (one-sided test for proportion P = 0.035). The enrichment of drug targets for paternal eGenes echoes the enrichment of GWAS hits for paternal eQTL (Table 2). This asymmetry between paternal and maternal eGenes perhaps finds its root in the conflicting interests of two sets of genes in relation to transfer of nutrients from the mother to her offspring (Moore and Haig, 1991). It is known that paternal alleles may have distinct evolutionary and functional roles compared to maternal alleles. Maze serves as an extreme example, where gene expression of the hybrid is regulated exclusively by the paternally transmitted alleles (Swanson-Wagner et al., 2009).

Discussion

Our study provides a comprehensive analysis of POE eQTL in human gene expression using data from the Framingham Heart Study. The extensive presence and complexity of POE uncovered in this study have significant implications for our understanding of genetic regulation and its role in complex traits and diseases. To illustrate the implications in the context of existing research, we revisit two previous findings in light of our new results, and advise caution on conducting Mendelian randomization using gene expression as the exposure.

T2D and maternal allele of rs2237892

SNP rs2237892 in the last intron of KCNQ1 was found in association with type 2 diabetes (T2D) in a Japanese cohort (Yasuda et al., 2008), and the association was later confirmed in a Chinese cohort (Liu et al., 2009). Three other SNPs (rs2283228, rs2237895, and rs2237897) in the same last intron of KCNQ1 were found in association with T2D in another Japanese cohort (Unoki et al., 2008). Both rs2237895 and rs2237897 were replicated in a Singaporean cohort, a Danish cohort (Unoki et al., 2008), and a Chinese cohort (Liu et al., 2009). In these studies, KCNQ1 was identified as T2D candidate gene.

Based on these results from the genotype test, Kong et al. (2009) demonstrated that SNP rs2237892 was maternally associated with T2D in an Iceland cohort. This maternal association was replicated with a larger odds ratio in a Pima Amerindian cohort (Hanson et al., 2013). According to data from GTEx, none of these SNPs are eQTL of KCNQ1 and CDKN1C in any tissue, consistent with the fact that these are imprinted genes and GTEx study is oblivious to parent of origin. Our eQTL analysis shows that SNPs rs2237892, rs2283228, and rs2237897 are maternal eQTL of CDKN1C, a down-stream neighbor of KCNQ1, and none of these SNPs are eQTL of KCNQ1.

We therefore suggest that CDKN1C, instead of KCNQ1, is a T2D candidate gene, for the following reasons: 1) These GWAS SNPs connect T2D and CDKN1C quantitatively, but connect T2D and KCNQ1 only geographically; 2) A study suggests that CDKN1C mutations may represent a novel monogenic form of diabetes (Kerns et al., 2014); 3) A boy carrying a frameshift mutation in CDKN1C was diabetic from week 29 (Berland et al., 2022); and 4) Targeted demethylation at the CDKN1C/p57 locus induces human β cell replication, while the loss of insulin-secreting β cell is characteristic among T1D and T2D (Ou et al., 2019).

Polar overdominance and rs1802710

Overdominance is a pattern of inheritance such that both homozygous AA and BB have the same baseline trait value, while heterozygous AB has a higher trait value. Polar over-dominance (Cockett et al., 1996) introduces asymmetry into overdominance to separates AB into two groups according to parent-of-origin of A (or B), such that AB with paternal A (or B) has a high trait value, while AB with maternal A (or B) has the baseline trait value.

Wermter et al. (2008) studied trios of extremely obese offspring and identified rs1802710 in exon 5 of DLK1, homologous to the ovine callipyge locus (Cockett et al., 1996), whose allelic transmission pattern was consistent with polar overdominance: frequent transmission of the paternal C allele to obese children, but the relative risk for carriers of the homozygous CC genotype was not increased compared to the reference TT genotype.

Our eQTL results show that DLK1 has no significant eQTL, but the maternal C allele of rs1802710 reduces expression of MEG3, 53Kb downstream of DLK1. Both MEG3 and DLK1 are imprinted with MEG3 maternally expressed and DLK1 paternally expressed. The expression of MEG3 was shown to be significantly higher in the obese group (Daneshmoghadam et al., 2021). Therefore the maternal C allele is associated with reduced risks of obesity, which balances out the increased risk conferred by paternal C allele, thus offering an explanation for the baseline risk of homozygous CC samples in (Wermter et al., 2008). Since rs1802710 is not an eQTL of any gene in any tissue according to GTEx data, our study is critical to link rs1802710 with MEG3, albeit not in the adipose tissue.

Mendelian randomization

Mendelian randomization (MR) refers to the random allocation of alleles at the time of gamete formation. Observational epidemiology studies use MR to infer the causal effect of an exposure on an phenotype (Smith and Ebrahim, 2003), as if the random allocation of alleles is comparable to a randomized clinical trial. An important assumption for MR, among many other important assumptions, is that the phenotype conferred by a specific genetic variant is homogeneous in the population, and exchangeable between paternally and maternally inherited alleles. POE is a direct violation of this assumption (Bochud et al., 2008).

There is a growing interest in using gene expression as an exposure to perform MR across various traits (Porcu et al., 2019; van der Graaf et al., 2020). If an exposure has eQTL with POE, ignoring the parent-of-origin will bias prediction of a subset of samples. If the bias is severe it will change the ranking of the exposure. Consequently, it either compromises the power of the MR or leads to a false conclusion of a causal relationship between the exposure and the phenotype.

Our study demonstrated the abundance of POE in gene expressions, and therefore advised caution when conducting MR using gene expression as the exposure. Our suggestion is checking the list of POE SNPs and eGenes we provided in the Supplementary Material and excluding those that show POE towards the exposure. The same suggestion is also applicable to computing polygenic risk scores, and imputing gene expression such as PrediXcan (Gamazon et al., 2015) and a Bayesian method motivated by it (Qi et al., 2018).

Limitations of the Study

Despite the significant findings, our study has several limitations. First, the analysis is restricted to whole blood gene expression, which may not capture POE in other tissues. Gene expression and POE can be tissue-specific, and further studies are needed to investigate POE across different tissues to gain a comprehensive understanding of its impact.

Second, the study relies on the availability of trios with both whole genome sequencing and RNA sequencing data, which limits the sample size. Although the Framingham Heart Study provides a relatively large number of trios, increasing the sample size and including more diverse populations would enhance the generalizability of our findings.

Finally, the functional implications of the identified POE eQTL require further validation through experimental studies. While our study provides a comprehensive catalog of POE eQTL, understanding their biological significance and mechanisms of action will require detailed experimental investigations.

Conclusion

Our study highlights the abundance and the complexity of POE in gene expression, providing a valuable resource that complements existing databases such as GTEx. These findings have significant implications for genetic association studies, the interpretation of complext traits, and the development of targeted therapies. By expanding the understanding of POE, this study contributes to a deeper comprehension of genetic regulation and its impact on human health.

Methods

Genotype data

We first used pedigree and availability of RNAseq data to narrow the samples down to 2955 individuals and identified all bi-allelic SNPs following a TOPMed protocol (Taliun et al., 2021). We then used Kindred to infer pairwise kinship for each chromosome, and computed mean and standard deviation (SD) of the kinship. A sample pair that is parent-offspring can be distinguished from full sibs by the SD (Supplementary Figure S2). Trios are consisted of sample A, B, and C such that ϕ(A, B) ≈ 0 and ϕ(A, C) ≈ ϕ(B, C) ≈ 0.25 where ϕ(·, ·) is the kinship. In the end, we identified 1477 trios that have whole genome sequencing data and whose children have RNAseq data. In total, we had 7, 752, 281 autosomal biallelic SNPs with minor allele frequency > 0.01 (not corrected for relatedness). Across all 1477 trios, the Mendelian error is negligible 4.76 per 100, 000 SNPs per trio. To handle Mendelian errors, we assumed that the child was correctly genotyped, and errors occurred in the parents, but if the child is heterozygous, we assumed this to be triple heterozygous SNP. On average 1477 trios had 65 triple heterozygous genotypes per SNP. The triple heterozygous SNPs cannot be phased by Mendelian inheritance, but can be phased based on a linkage disequilibrium model (below).

Inference of parent of origin

For each trio, we first phased all markers that are not triple heterozygous using rules of Mendelian inheritance, where for the child in each trio we obtained paternal and maternal haplotypes with triple heterozygous markers whose phase were unresolved. We marked those triple heterozygous markers as missing. For markers that showed Mendelian incompatibility, we assumed that the child’s genotype was correct and marked it as missing if it was heterozygous. Using a hidden Markov model that was designed to model haplo-type variation (Guan, 2014), we imputed missing markers in each haplotypes to obtain dosage estimates (between 0 and 1). Then for each marker that was marked as missing, we compared imputed dosages between paternal and maternal haplotypes, and assigned the large dosage as 1 and small dosage as 0. We thus obtained paternal and maternal haplo-types with no missing data. We simulated trios using haplotypes from the 1000 Genomes project (Auton et al., 2015), and investigated the accuracy of the phasing by the mask and imputation approach described above. The phasing was highly accurate, the error rate was less than 1 out of 200 triple heterozygous SNPs (Supplementary Table S1).

RNAseq data

The RNAseq data of 1477 children was obtained in three batches with 1381 in batch 1; 14 in batch 2, and 82 in batch 3. Our primary goal was to correct for GC content bias. To this end, we first removed genes whose proportion of 0 TPM values was greater than 5%, which retained 16, 969 genes out of total 58103, a majority of which are coding genes. Among those, 16, 824 genes have GC content. To correct for GC content bias, we took an approach used by EDAseq to regress out the GC content from log(1+TMP), but instead of using local linear regression as documented in EDAseq, we used local quadratic regression implemented in loess in R (Supplementary Figure S3). Specifically, let y = log(1 + TMP) be a gene expression and g be the GC content, we fit ly = loess(y, g) and compute ŷ = median(y) + ly$residuals. We then quantile normalized ŷ by qqnorm(ŷ, plot.it = F)$y (in R) separately for each gene.

eQTL analysis

We used IDUL to perform eQTL analysis. IDUL fits linear mixed models to achieve exact optimal. It was specifically developed to analyze multi-omics data to achieve high efficiency by reusing the intermediate computations (Guan and Levy, 2024a). Previous analysis showed that for the Framingham Heart Study participants, the genomic inflation was well controlled using a linear mixed model with kinship matrix computed by Kindred (Guan and Levy, 2024c). In this study, we also controlled age, sex, BMI, and white blood cell compositions in peripheral blood. We now briefly describe how Bayes factors were computed. Consider a model where W contains conventional covariates such as age and sex, including a column of 1, x contains genetic variant(s) to be tested for association, u is the random effect with Z as its loading matrix and twice of kinship K as its covariance (both Z and K are known), MVNn denotes an n-dimensional multivariate normal distribution, In is n-dimensional identity matrix. Denote X = (W, x) and b = (a, β), then Xb is the fixed effect, and we assume X has a full rank c. In genetic association studies, the random effect Zu is a nuisance term that absorbs part of the phenotype y that is attributable to population stratification and relatedness. The maximum likelihood estimate (MLE) of η can be efficiently obtained (Guan and Levy, 2024a), plug back into (1), and specify the following conjugate prior and let Va → ∞, κ1 → 0 and κ2 → 0, Bayes factor can be evaluated efficiently in a closed form. In this study we used σ = 0.5, following prior work of Bayes factors for linear models (Servin and Stephens, 2007). Complete details on computing Bayes factors for linear mixed model can be found in (Guan and Levy, 2024b).

P-value for difference between paternal maternal effects

For joint analysis, we have with the prior for β1 and β2 as The Bayes factor can be similarly computed in a closed form. We obtain the posterior estimates of and , their variances and and covariance s12. To test null hypothesis , we compute test statistics . Since under the null t following χ2 distribution with 1 degree of freedom, we can compute a p-value.

Threshold for eQTL sets and gene sets

For paternal eQTL set SP, we required log10 BF1 > 4 and log10 BF0 < θ; For maternal eQTL set SM, we required log10 BF0 > 4 and log10 BF1 < θ; For opposing eQTL set SO, we required log10 BFj − log10 BFg > 4 and log10 BF1 > θ and log10 BF0 > θ, and β1 ∗ β0 < 0; For genotype eQTL set SG, we required log10 BFg > 4 and log10 BF1 > θ and log10 BF0 > θ, and β1 ∗ β0 > 0. It’s easy to see for a particular type of eQTL, different θ produced nested eQTL sets. We tried θ = 0, log10 2, and log10 3 to obtain eQTL sets (Supplementary Table S2), from which we obtained gene sets harboring those eQTL GP, GM, GA and GG, from which we obtained genes harboring exclusively paternal eQTL G1 = GP \ (GM ∪ GA ∪ GG), genes harboring exclusive maternal eQTL G0 = GM \ (GP ∪ GA ∪ GG), and genes harboring exclusively opposing eQTL G0 = GA \(GP ∪ GM ∪ GG). We compared known imprinted genes with paternal expression with G1, and known imprinted genes with maternal expression with G0. θ = log10 2 produced minimum G1 and G0 that contain known imprinted genes.

Availability of data and materials

This study analyzed existing datasets from public domain and simulated datasets. Data from the 1000 Genomes project can be found at https://www.internationalgenome.org, and datasets from Framingham Heart Study can be obtained via dbGaP. The pipeline and software used to conduct analysis is available at https://github.com/haplotype/poe. The results (POE eQTLs) generated from current study will be made available to general public once the manuscript is in print.

eLife Assessment

This important study provides a systematic investigation of parent-of-origin (POE) effects on gene expression using large trio-based data from the Framingham Heart Study, uncovering thousands of potentially novel associations. While the findings are potentially significant, the statistical support for classifying POE eQTLs and some downstream analyses is incomplete, and more stringent re-analysis is needed. With such revisions, the work would serve as a foundation for advancing understanding of POEs and their role in gene regulation.

Reviewer #1 (Public review):

Summary:

This study presents a systematic investigation of parent-of-origin effects on gene expression using trio-based data from the Framingham Heart Study, which is notable for its relatively large number of trios. By combining whole-genome and RNA sequencing data, the authors examined the extent to which gene expression is influenced by whether genetic variants are inherited maternally or paternally.

The authors report that parent-of-origin eQTLs are widespread, identifying 15,893 eQTLs from 14,733 variants and 1,824 genes that were significant in paternal, maternal, or joint tests but not detected by traditional eQTL approaches. They further classified these associations based on the relative strength and direction of paternal and maternal effects, highlighting a subset with opposing directions. The study also highlighted eGenes linked to known imprinted genes as well as those with opposing parent-specific effects, and observed that paternal eGenes are enriched for drug targets. Finally, the work revisits previous findings in which eQTL studies were used to interpret disease-associated loci, emphasizing that conventional eQTL analyses without testing the parent-of-origin may mislead gene prioritization efforts. The study recommends that future downstream analyses, such as Mendelian randomization, take into account the provided lists of SNPs and eGenes and exclude those with strong parent-of-origin effects when linking genetic regulation to disease risk.

Strengths:

The major strength of the study lies in the scale and quality of the dataset, the trio-based design, and the systematic application of statistical tests for parent-of-origin effects. The strengths thoughtfully employed Bayes factors rather than p-values to provide stronger evidence of association, which adds rigor to their analyses. These design choices provide compelling evidence that parent-of-origin effects are widespread and that conventional eQTL analyses miss a substantial fraction of regulatory variation. The results are clearly presented and supported by robust analyses, including the identification of opposing parental effects and the enrichment of paternal eGenes for drug targets. Notably, the two examples demonstrating how these findings can reshape disease gene prioritization highlight the broader impact of the study and encourage further work in the community to incorporate parent-of-origin effects.

Weaknesses:

The main limitations of the study are threefold. First, there is a lack of replication in independent cohorts, which is understandable given the difficulty of identifying datasets with a comparable number of trios, but replication would help establish the generalizability of the findings. Second, while Bayes factors are thoughtfully used to assess evidence of association, the paper does not fully explore how the chosen thresholds translate to the expected rate of false positives. For example, a minor allele frequency cutoff of 1% was applied, which seems somewhat arbitrary, and without reporting the allele frequency distribution of the identified eQTLs, it is unclear whether rare variants disproportionately contribute to the signals, potentially affecting the reliability of discoveries. Third, the ancestry background of the study samples is not reported, which could be a confounding factor in the genetic analyses.

Reviewer #2 (Public review):

Summary:

The authors have used 1477 sequenced trios with available gene expression data in the offspring to discover eQTLs that act in a parent-of-origin specific manner. The classified associated SNPs are tested for enrichment for GWAS hits, drug target genes, etc.

Strengths:

The manuscript presents an impressive analysis of a very rich data set of parent-of-origin eQTLs. To my knowledge, it is one of the largest studies of its kind, most analyses are sound, and the results are of interest to many in the field and potentially beyond. The different ideas of follow-up analyses are useful and make sense.

Weaknesses:

While in general the analyses are well-conducted, I noticed a major issue with the POE eQTL classification, which puts into question most of the downstream analysis. In light of this problem, most of the analysis would need to be rerun, which represents a major revision of the paper, but is straightforward to repair.

The major problem with the classification of POEs is that simply having significant maternal, but insignificant paternal effect is not an indicator of POE, this happens widely for SNPs with no POE whatsoever (it can happen by chance even when both maternal and paternal effects are the same and non-zero - the authors can see it via simulations under the null [maternal=paternal effect]). In order to be able to talk about POE, first, a significant difference between maternal and paternal effects needs to be claimed. Therefore, none of the 4 sets of POE eQTLs are justified. To me, the only relevant criterion to pick POE SNPs is the P-value when comparing the maternal and paternal effects. The definitions of the 4 groups are based on somewhat ad hoc priors, BF thresholds, etc. Also, in Section 4.6, the value of theta is arbitrarily chosen (along with the threshold of 4 to declare POE). In my opinion, the clean treatment of the 4 groups would start with a significant P-value (beta_maternal vs beta_paternal). Within this set, you can then use the original criteria presented in the paper, but only among these associations where there is solid evidence of different parental effects.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer 1 (Public review):

Summary:

This study presents a systematic investigation of parent-of-origin effects on gene expression using trio-based data from the Framingham Heart Study, which is notable for its relatively large number of trios. By combining whole-genome and RNA sequencing data, the authors examined the extent to which gene expression is influenced by whether genetic variants are inherited maternally or paternally.

The authors report that parent-of-origin eQTLs are widespread, identifying 15,893 eQTLs from 14,733 variants and 1,824 genes that were significant in paternal, maternal, or joint tests but not detected by traditional eQTL approaches. They further classified these associations based on the relative strength and direction of paternal and maternal effects, highlighting a subset with opposing directions. The study also highlighted eGenes linked to known imprinted genes as well as those with opposing parent-specific effects, and observed that paternal eGenes are enriched for drug targets. Finally, the work revisits previous findings in which eQTL studies were used to interpret disease-associated loci, emphasizing that conventional eQTL analyses without testing the parent-of-origin may mislead gene prioritization efforts. The study recommends that future downstream analyses, such as Mendelian randomization, take into account the provided lists of SNPs and eGenes and exclude those with strong parent-of-origin effects when linking genetic regulation to disease risk.

Strengths:

The major strength of the study lies in the scale and quality of the dataset, the trio-based design, and the systematic application of statistical tests for parent-of-origin effects. The strengths thoughtfully employed Bayes factors rather than p-values to provide stronger evidence of association, which adds rigor to their analyses. These design choices provide compelling evidence that parent-of-origin effects are widespread and that conventional eQTL analyses miss a substantial fraction of regulatory variation. The results are clearly presented and supported by robust analyses, including the identification of opposing parental effects and the enrichment of paternal eGenes for drug targets. Notably, the two examples demonstrating how these findings can reshape disease gene prioritization highlight the broader impact of the study and encourage further work in the community to incorporate parent-of-origin effects.

Weaknesses:

The main limitations of the study are threefold.

First, there is a lack of replication in independent cohorts, which is understandable given the difficulty of identifying datasets with a comparable number of trios, but replication would help establish the generalizability of the findings.

We fully agree with the reviewer that replication in an independent cohort is a crucial step for establishing generalizability. As the reviewer notes, the Framingham Heart Study, with its 1,477 trios possessing both WGS and RNA-seq data, represents a uniquely powerful and, to our knowledge, currently unmatched resource for this specific type of parent-of-origin eQTL analysis.

In the absence of an external cohort of comparable size and data richness, we have taken several steps to ensure the internal validity and robustness of our findings within the current study, which we will clarify and expand upon in the revised manuscript:

Positive Control Validation: We explicitly used well-established, bona fide imprinted genes (e.g., MEG3, NDN, SNURF, as listed in Table 1 and Figure 1) as positive controls. The fact that our analysis correctly identifies their known parent-of-origin expression patterns (e.g., maternal eQTL for MEG3, paternal eQTL for NDN) serves as a powerful internal validation of our phasing methodology, statistical models, and significance thresholds. This demonstrates that our approach has the power to detect true POE signals.

Conservative Calling Criteria: As the reviewer suggests, we prioritized specificity. Our definition of eQTL sets (Section 4.6) uses stringent thresholds (e.g., log10 BF > 4 for primary signals and θ = log10 2 for exclusivity). We explored different θ parameters (Supplementary Table S2) and chose the one that minimized the inclusion of false positives, ensuring that our core gene sets (e.g., G1,G0,G2) are high-confidence discoveries.

Rigorous Analytical Pipeline: As we note in the revised text, our conclusions are supported by a robust analytical pipeline. This includes trio-based phasing validated by simulation (Supplementary Table S1), the use of linear mixed models to control for relatedness and population structure, and the application of Bayes factors which inherently penalize variants with low minor allele frequencies, thereby reducing spurious associations.

We believe these internal consistency checks and methodological rigor provide strong confidence in our findings. To further facilitate external replication, we will make the full list of POE eQTLs and eGenes available as a comprehensive resource (as noted in the Discussion and Supplementary Materials), enabling other researchers to validate these findings as appropriate datasets become available.

Second, while Bayes factors are thoughtfully used to assess evidence of association, the paper does not fully explore how the chosen thresholds translate to the expected rate of false positives. For example, a minor allele frequency cutoff of 1% was applied, which seems somewhat arbitrary, and without reporting the allele frequency distribution of the identified eQTLs, it is unclear whether rare variants disproportionately contribute to the signals, potentially affecting the reliability of discoveries.

We thank the reviewer for raising this important point regarding the calibration of our significance thresholds and the potential role of rare variants. We address this by clarifying the relationship between Bayes factors, prior odds, and false discovery rates, and by providing a more detailed characterization of the variants we identified.

Bayes Factors and False Discovery: The reviewer is correct that the connection between a Bayes factor threshold and a false positive rate is not direct as it has to take into account of prior odds. As we briefly noted, for a given prior odds of association (e.g., 1 in 100 or 1 in 1000 for a cis-eQTL), a log10 BF = 4 corresponds to a posterior probability of association (PPA) of 0.99 or 0.90 respectively. Consequently, 1 − PPA can be interpreted as the local false discovery rate (lfdr), as we have now explicitly stated in Section 2.2 (citing Soloff et al., 2024). Our choice of log10 BF = 4 was therefore chosen to ensure a very low or modest lfdr (depending on the prior odds) for our primary findings.

Minor Allele Frequency Threshold: The 1% MAF cutoff was indeed a pre-analysis filtering step. It was chosen based on the power afforded by our sample size of 1,477 trios. For variants rarer than 1%, our study is underpowered to detect associations, and any signals would be highly unstable. Importantly, the reviewer’s concern about rare variants disproportionately contributing to signals is further mitigated by our use of Bayes factors. As we note in Section 2.2, the prior used in our Bayes factor computation (with σ = 0.5 in the prior for effect sizes, as described in Section 4.4) inherently penalizes variants with small minor allele frequencies. This is because for a given effect size, the evidence for association is weaker for a rare variant than a common one. Thus, the combination of a pre-analysis MAF filter and the Bayesian analysis itself guards against spurious findings driven by very rare alleles.

Allele Frequency Distribution: To directly address the reviewer’s request for transparency, in the revised manuscript we include a supplementary figure (e.g., Supplementary Figure S4) showing the distribution of minor allele frequencies (1000 genomes European descents) for the SNPs identified in paternal eQTL set SP and maternal eQTL set SM. This empirically demonstrate that our findings are not disproportionately driven by low-frequency variants and provide a more complete picture of the genetic architecture underlying these POE signals. We also add a sentence to the Results section (Section 2.5) summarizing this distribution.

Third, the ancestry background of the study samples is not reported, which could be a confounding factor in the genetic analyses.

We thank the reviewer for highlighting this omission. In the revised manuscript, we explicitly report the ancestry background of the Framingham Heart Study participants analyzed. Consistent with previous reports on this cohort, the vast majority of samples are of European descent.

Crucially, as the reviewer suggests, population stratification can be a confounder in genetic studies. To mitigate this, our analysis employed a linear mixed model (Section 4.4) that includes a random effect with a covariance structure defined by the genetic relatedness matrix (GRM). This approach is specifically designed to control for spurious associations due to both subtle population structure and known relatedness among individuals, ensuring that our findings are robust to these potential confounders.

Reviewer 2 (Public review):

Summary:

The authors have used 1477 sequenced trios with available gene expression data in the offspring to discover eQTLs that act in a parent-of-origin specific manner. The classified associated SNPs are tested for enrichment for GWAS hits, drug target genes, etc.

Strengths:

The manuscript presents an impressive analysis of a very rich data set of parent-of-origin eQTLs. To my knowledge, it is one of the largest studies of its kind, most analyses are sound, and the results are of interest to many in the field and potentially beyond. The different ideas of follow-up analyses are useful and make sense.

Weaknesses:

While in general the analyses are well-conducted, I noticed a major issue with the POE eQTL classification, which puts into question most of the downstream analysis. In light of this problem, most of the analysis would need to be rerun, which represents a major revision of the paper, but is straightforward to repair.

We appreciate the reviewer’s concern and take it seriously. However, we believe the issue stems from a misunderstanding of our classification framework. We clarify our reasoning below, and we are confident that no re-analysis is necessary. In fact, our Bayesian approach was specifically chosen to avoid the very problem the reviewer raises.

The major problem with the classification of POEs is that simply having significant maternal, but insignificant paternal effect is not an indicator of POE, this happens widely for SNPs with no POE whatsoever (it can happen by chance even when both maternal and paternal effects are the same and non-zero - the authors can see it via simulations under the null [maternal=paternal effect]).

The reviewer raises a valid statistical concern: under the null hypothesis of equal maternal and paternal effects (β0 = β1≠ 0), sampling variation could occasionally produce a scenario where one effect appears significant and the other does not. This is indeed a form of Type II error (failing to detect a true non-zero effect for one of the alleles).

However, this is precisely why we chose Bayes factors over p-values. A key advantage of Bayes factors is that they are not blind to power. P-values are calculated solely under the null hypothesis and do not incorporate any information about the alternative hypothesis or the study’s power to detect it. Consequently, when power is low (e.g., due to minor allele frequency differences between paternal and maternal alleles), p-values can be misleading.

In contrast, Bayes factors are computed under both the null and alternative hypotheses. They inherently incorporate power through the prior specification. As we note in Section 2.2, “Bayes factors penalize genetic variants with small allele frequencies to reduce false positives.” This means that a SNP where, by chance, one allele appears significant and the other does not—but where power is low due to allele frequency imbalance—will not receive a high Bayes factor, because the evidence is appropriately discounted.

In order to be able to talk about POE, first, a significant difference between maternal and paternal effects needs to be claimed. Therefore, none of the 4 sets of POE eQTLs are justified. To me, the only relevant criterion to pick POE SNPs is the P-value when comparing the maternal and paternal effects.

We respectfully disagree with the reviewer’s assertion that our approach to POE eQTL classification are not justified. There are multiple biologically meaningful patterns of parent-of-origin effects, and our classification scheme was designed to capture this diversity:

(1) Paternal-specific eQTL (β0 = 0, β1 ≠ 0)

(2) Maternal-specific eQTL (β0 ≠ 0, β1 = 0)

(3) Opposing eQTL (β0 ≠ 0, β1 ≠ 0,β0 × β1 < 0)

(4) Genotype eQTL (β0= β1 ≠ 0)

The reviewer’s proposed test (H0: β0 = β1) collapses these distinct biological scenarios into a single binary outcome. For example: A purely paternal-specific eQTL (β0 = 0, β1 ≠ 0) would indeed show a significant difference, and would be captured by the reviewer’s test. However, a gene like ZNF890P in Table 1, where both effects are significant and in the same direction but of different magnitudes, would also show a significant difference. In the reviewer’s framework, this would be classified as a POE eQTL, yet biologically it behaves more like a genotype eQTL with an allelic imbalance. Our framework correctly separates these cases.

Moreover, the reviewer’s proposed test is a nested special case of our broader approach. As we note in our response, our paternal-specific test (H0: β0 = β1 = 0 vs H1: β0 = 0,β1 ≠ 0) is a more constrained hypothesis that yields a subset of the SNPs that would be identified by the reviewer’s difference test, were it to have sufficient power. Our approach is therefore more conservative for classifying paternal- or maternal-specific eQTLs, not less.

The definitions of the 4 groups are based on somewhat ad hoc priors, BF thresholds, etc. Also, in Section 4.6, the value of theta is arbitrarily chosen (along with the threshold of 4 to declare POE). In my opinion, the clean treatment of the 4 groups would start with a significant P-value (beta-maternal vs beta-paternal). Within this set, you can then use the original criteria presented in the paper, but only among these associations where there is solid evidence of different parental effects.

We take strong issue with the characterization of our prior specifications and thresholds as “ad hoc” or “arbitrary.” In Bayesian analysis, prior specification is a principled and transparent modeling choice, not an arbitrary one.

(1) Choice of log10 BF = 4 threshold: As stated in Section 2.2, this threshold was chosen based on explicit considerations of prior odds and posterior probability of association. For a prior odds of 1:1000 (a reasonable guess for cis-eQTLs), this BF corresponds to a posterior probability of association of 0.91. If one prefers a more optimistic prior odds of 1:100, the PPA becomes 0.99. The threshold is therefore grounded in decision theory, not whim.

(2) Choice of θ in Section 4.6: We explicitly state that we explored multiple values of θ(0, log10 2, log10 3) and chose θ = log10 2 because it “produced minimum G1 and G0 that contain known imprinted genes.” This is a principled, data-driven calibration step using positive controls, not an arbitrary selection. The transparency of this process is a strength, not a weakness.

(3) Comparison to p-value thresholds: The reviewer suggests that p-value thresholds are somehow less arbitrary. However, the conventional p-value threshold of 0.05 is itself a historical convention with no universal justification. Moreover, as we note, p-values do not account for power differences across SNPs. A p-value of 5 × 10−8 from a SNP with 40% MAF is not comparable to the same p-value from a SNP with 1% MAF, because the power to detect the association differs dramatically. Bayes factors automatically adjust for this through the prior, making them more comparable across variants, not less.

In revision, we added a section in supplementary to review relationships between p-values, Bayes factors, and FDR.

Recommendations for the authors:

Reviewer 1 (Recommendations for the authors):

Here are some suggestions to improve the study:

(1) Provide information about the ancestry background of participants and consider including ancestry principal components in the eQTL models, as is commonly done, to account for population structure.

We thank the reviewer for this suggestion. In the revised manuscript, we explicitly state that the participants in the Framingham Heart Study are predominantly of European descent, consistent with previous publications from this cohort. Regarding population structure, we respectfully note that our analysis already employs a linear mixed model (Section 4.4) that includes a random effect with a covariance structure defined by the genetic relatedness matrix (GRM). This approach is widely regarded as more robust than including a limited number of principal components, as it accounts for both fine-scale population stratification and known relatedness simultaneously.

(2) Conduct sensitivity analyses using different Bayes factor cutoffs to assess the robustness of the findings.

We appreciate the reviewer’s concern about threshold robustness. In fact, we already conducted a form of sensitivity analysis during the classification step. As described in Section 4.6 and shown in Supplementary Table S2, we explored multiple values of θ (0, log10 2, and log10 3) and observed how they affected the composition of our gene sets. The choice of log10 BF = 4 for significance was similarly grounded in posterior probability calculations (Section 2.2). To further address the reviewer’s point, we add a Supplementary Table S3 for counts of eQTL and eGenes under different Bayes factor threshold. This demonstrates that our most significant claim, the abundance of POE eQTL, are not overly sensitive to the specific cutoff.

(3) In the GWAS examples for KCNQ1 and CDKN1C, the assessment of whether the SNPs act as eQTLs for the two genes is based on a single BF threshold, which may be influenced by differences in gene expression levels. The authors could compare the corresponding effect sizes of these SNPs on both genes to provide a more nuanced investigation. While the limitation of missing data from other tissues is discussed in the paper, it remains possible that KCNQ1 plays a role in tissues more relevant to T2D.

This is an excellent suggestion for a more nuanced investigation. We re-examined the effect sizes for the SNP rs2237892 in our published results. For gene CDKN1C, the paternal log10 BF1 = −0.477 and maternal log10 BF0 = 4.94, the normalized maternal effect in joint analysis is −4.86 vs −0.74 for paternal. Unfortunately, the published results has no eQTL for KCNQ1, which according to our selection creteria means maximum log10 BF < 3 for all tests (genotype, paternal , maternal, joint). The concern for different gene expression level may affect BF is valid. We preempt this pitfall by quantile normalization of gene expression levels after controlling for GC content (as documented in Method Section). We agree with the reviewer that the lack of data from pancreatic tissues is a limitation. We add a sentence in revelant section to acknowledging that while whole blood is a valuable and accessible tissue, replication in T2D-relevant tissues (e.g., pancreas, adipose) would be an important future direction, and our findings provide a hypothesis for such targeted investigations.

Reviewer 2 (Recommendations for the authors):

Major comments:

There are some literature elements missing:

(1) Hofmeister has a newer and larger study [https://pubmed.ncbi.nlm.nih.gov/40770099/].Please cite that too; it also has POE pQTLs, which is relevant.

(2) POE in pigs has been explored [https://www.nature.com/articles/s41467-02562243-6], please cite it.

(3) An insightful review covering the mechanisms of POE for gene expression (https://www.sciencedirect.com/science/article/pii/S2352154618300482) should be cited.

(4) Further studies on POE in gene expression in social insects (https://royalsocietypublishing.org and in mice (https://www.biorxiv.org/content/10.1101/2023.08.24.554674v1.full) are also relevant.

We thank the reviewer for bringing these important references to our attention. We incorporated the suggested citations in the revision to provide a more comprehensive context for our work, including the newer POE pQTL study by Hofmeister et al., the findings in pigs, and the mechanistic review.

While it’s OK to report and rank SNPs by BF, it is necessary to show association P-values as well. It is not explained in the text around the Table how the P-value is obtained in the Table. And it is important to show how their priors translate to FWER control. What is the FWER when picking SNPs at a certain BF value? 1-PPA and local FDR depend on the choice of the prior, but we need a prior-independent measure of FDR/FWER.

We appreciate the opportunity to clarify. The p-value presented in Table 1 (column “P”) is indeed the frequentist p-value testing the null hypothesis of equal maternal and paternal effects (H0 : β0 = β1), as described in Section 4.5. We included this to provide a familiar metric for readers, but our discovery framework relies on Bayes factors for the reasons outlined in Section 2.2.

Regarding error control, the reviewer is correct that 1-PPA is a local FDR that depends on the prior. We chose to control the local rate of false discoveries rather than the Family-Wise Error Rate (FWER) because FWER control (e.g., via Bonferroni) is often excessively conservative for exploratory analyses like eQTL mapping, especially given the correlation among tests due to LD.

Our Bayesian approach provides a more nuanced measure of evidence at the level of each individual test, which is precisely what is needed for prioritizing SNPs with parent-of-origin effects.

The demand for a prior-independent measure of FDR is conceptually problematic. Any probabilistic statement about a specific hypothesis being true or false necessarily requires a prior—this is a fundamental consequence of probability theory. Frequentist FDR, while prior-independent in one sense, does not provide a probability that a particular finding is false; it is a long-run error rate over many tests. Methods like q-values, often described as “prior-free,” still depend on implicit assumptions (e.g., the estimate of π0, independence of tests, and a mixture of effect sizes).

In our specific context of cis-eQTL analysis, these assumptions are particularly questionable. LD induces correlation among nearby SNPs, violating the independence required for stable π0 estimation. Moreover, effect sizes in a region are not randomly mixed—SNPs in high LD tend to have similar effect directions and magnitudes, which can bias the mixture model underlying q-value approaches. Our Bayesian approach, by modeling each SNP individually, avoids these cross-SNP assumptions.

Importantly, while posterior probabilities depend on the choice of prior (π0), we have verified that our conclusions are robust across a wide range of plausible π0 values (0.9,0.99,0.999). Given our extremely stringent Bayes factor threshold (BFj > 104), the posterior probability for a maternal effect exceeds 0.90 for any π0 < 0.999. Thus, the prior dependence is practically irrelevant for the SNPs we report.

In revision, we added a section in Supplementary to describe the connections between p-value, Bayes factor, and FDR. We hope this will clarify that a (seemingly) prior independent FDR has a hidden assumption that cis-eQTL analysis is likely to violate.

The major problem with the classification of POEs is that simply having significant maternal, but insignificant paternal effect is not an indicator of POE, this happens widely for SNPs with no POE whatsoever (it can happen by chance even when both maternal and paternal effects are the same and non-zero - the authors can see it via simulations under the null [maternal=paternal effect]). In order to be able to talk about POE, first, a significant difference between maternal and paternal effects needs to be claimed. Therefore, none of the 4 sets of POE eQTLs are justified. To me, the only relevant criterion to pick POE SNPs is the P-value when comparing the maternal and paternal effects. The definitions of the 4 groups are based on somewhat ad hoc priors, BF thresholds, etc. Also, in Section 4.6, the value of theta is arbitrarily chosen (along with the threshold of 4 to declare POE). In my opinion, the clean treatment of the 4 groups would start with a significant P-value (beta-maternal vs beta-paternal). Within this set, you can then use the original criteria presented in the paper, but only among these associations where there is solid evidence of different parental effects.

We respectfully disagree with the reviewer’s assertion that a significant difference between maternal and paternal effects is the only valid criterion for defining POE, and we maintain that our classification is statistically sound and biologically meaningful.

The Problem with the “Difference-Only” Approach: The reviewer’s proposed filter (a significant p-value for β0 ≠ β1) is a single hypothesis test. Our goal was to classify eQTLs into multiple, distinct biological categories (paternal-specific, maternal-specific, opposing, etc.). The “difference-only” test collapses these categories. For example, a purely paternal-specific eQTL (β0 = 0,β1 ≠ 0) and a gene like ZNF890P (β0 ≠ 0, β1 ≠ 0, β0 > β1) would both show a significant difference. In the reviewer’s framework, they would be lumped together, obscuring the fact that one is an imprinted gene and the other is a standard eQTL with allelic imbalance. Our framework correctly separates them.

Bayes Factors are Not “Ad Hoc”: The choice of prior (σ = 0.5) follows established literature for linear model Bayes factors (Servin and Stephens, 2007). The threshold of log10 BF = 4 was chosen based on its relationship to posterior probability (0.91-0.99 given reasonable prior odds), which is a transparent and principled decision rule. The selection of θ in Section 4.6 was calibrated using a positive control set of known imprinted genes, ensuring our definitions were conservative and accurate. This is the opposite of arbitrary.

The Suggested Procedure Has Low Power: One can run the following simple R code to verify. We simulate maternal alleles xx and maternal alleles yy, then simulate phenotype with βxx > 0 and βyy = 0 (maternal effect only). We fit the joint model and compute p-values for the null βxx = βyy as suggested by reviewer. From the joint fit, we also extract p-values based on the null βxx = 0 and βyy = 0 respectively. The simulation was repeated 1000 times and p-values were stored in a matrix.

We call positives based on suggested procedure, and compare number of positives called using marginal p-values at two threshold of 1×10−5 and 1×10−6 to declare significance. We used threshold of 0.01 to declare insignificance.

The result demonstrates that the suggested procedure has a much lower power compared to the procedure based on marginal statistics.

For the above reasons, the follow-up enrichment analysis is somewhat questionable. Most enrichments are non-significant, and it is likely because the SP and SM groups are diluted with SG SNPs. The P1-P9 groups have nothing to do with POE, and although the observation of increased enrichment for GWAS SNPs with increased pleiotropy is interesting, it is irrelevant for POE.

We will address the dilution concern below. We agree that P1-P9 groups are not directly related to POE. But this is an interesting observation non-theless. As we found such an observation is missing in the literature, we ask to keep it in the paper.

In the same way, section 2.7 is not supported; the claimed maternal and paternal POEs are heavily diluted by simple marginal associations. The same holds for sections 2.82.10. A striking example is Table 3: for clinical trial targets, paternal/maternal eQTLs behave just like simple marginal eQTLs (GG). A similar pattern emerges for combined target enrichment.

The reviewer’s concern that our SP and SM sets are “diluted with SG SNPs” is precisely the issue our Bayes factor thresholds were designed to prevent. By requiring one effect to be significant and the other to be below a low threshold (θ), we explicitly excluded SNPs where both effects are significant and in the same direction (which defines SG).

Regarding Table 3, the reviewer’s interpretation differs from ours. The fact that paternal eQTLs (GP) show significant enrichment for drug targets, while genotype eQTLs (GG) also show enrichment, does not imply dilution. Rather, it suggests there is an overlap in the biological importance of these gene sets, which is expected. The key message of the finding is the asymmetry: GP is significantly more enriched than GG (p=0.035 for combined targets), a pattern that would be washed out if GP were merely a diluted version of GG. This asymmetry supports the interesting biological hypothesis (Moore and Haig, 1991) we discuss. The non-significance for GM further highlights this asymmetry.

I’m not sure how MR would be biased by POE: MR is conducted only if there is a marginal association, i.e., the average maternal and paternal effects are significant. If the expression is causal for a trait, the POE effect is propagated to the outcome; hence, the SNP effect on the exposure will be equally biased as the SNP effect on the outcome, and these cancel out, and the causal effect remains unbiased. Can the authors propose a concrete example of maternal/paternal effects that demonstrates their claimed bias?

We thank the reviewer for this insightful question, which allows us to clarify our point with a concrete example from our data.

Consider a scenario where one wishes to use Mendelian Randomization (MR) to test whether the expression of gene NECAB3 causally influences a particular trait (e.g., obesity). The reviewer is correct that if the causal effect is homogeneous, the average effect might still be captured. However, the bias we caution against arises in stratified analyses or in the interpretation of the genetic instrument itself.

Take the SNP rs4911348 and its effect on NECAB3 (Figure 2). The genotype model shows no marginal association. Therefore, if a researcher were conducting a standard MR study using this SNP as an instrument for NECAB3 expression, they would discard it as an invalid instrument due to the lack of a marginal association. They would miss the true underlying biology entirely. The causal effect of NECAB3 on the trait would be masked in the full population.

More subtly, even if a SNP has a marginal association, using it as an instrument while ignoring POE can lead to incorrect effect estimates in population subgroups defined by parent of origin. This is analogous to ignoring effect modification. For instance, if a treatment (exposure) has a different effect depending on which parent it came from (which is impossible, but the genetic propensity for the exposure does), failing to account for this can bias the instrumental variable estimate if the instrument’s strength varies by an unmeasured factor (parental origin).

Our advice to “check the list of POE SNPs” is a practical caution: if the instrument for an exposure exhibits strong POE, the standard MR assumptions about the homogeneity of the instrument’s effect may be violated, potentially leading to biased estimates or incorrect conclusions about causality.

Minor comments:

(1) In Table 1, the last column header should be -log10(P), not ”P”.

The column labelling is an editorial choice to prevent table overflow. This particularly labelling was explained in the caption.

(2) While BFg/0/1/j are explained in the text, these notations should be explained in the Table caption as well.

Added explanation in caption.

(3) It should also be mentioned in the Table 1 caption how these top 10 SNPs were chosen.

These are sentinel eQTL for each gene. We think the first paragraph of Section 2.3 explains clearly.

(4) “may ”acquires” a cis-eQTL through” → ”may ”acquire” a cis-eQTL through”.

Corrected. Thank you.

(5) “which retained 16, 969 genes out of total 58103”, I assume the 58103 are transcripts, not genes.

You are absolutely correct. We added transcripts after 58103.

(6) In Equation (1), Z is not defined. In this concrete setting, isn’t it simply the identity matrix?

Yes. Z is the identitity (loading) matrix for human study. We added a sentence to clarify in revision.

Version 2

Abstract

Parent-of-origin effect (POE) is a phenomenon whereby an allele’s effect on a phenotype depends both on its allelic identity and parent’s gender from whom the allele is inherited, as exemplified by the polar overdominance in the ovine callypyge locus and the human obesity DLK1 locus. Systematic studies of POE of expression quantitative trait loci (eQTL) are lacking. In this study we use trios among participants in the Framingham Heart Study to examine to what extend POE exists for gene expression of whole blood using whole genome sequencing and RNA sequencing. For each gene and the SNPs in cis, we performed eQTL analysis using genotype, paternal, maternal, and joint models, where the genotype model enforces the identical effect sizes on paternal and maternal alleles, and the joint model allows them to have different effect sizes. We compared models using Bayes factors to identify paternal, maternal, and opposing eQTL, where paternal and maternal effects have opposite directions. The resultant variants are collectively called POE eQTL. The highlights of our study include: 1) There are more than 2, 000 genes harbor POE eQTL and majority POE eQTL are not in the vicinity of known imprinted genes; 2) Among 180 genes harboring opposing eQTL, 99 harbor exclusively opposing eQTL, and 58 of the 99 are phosphoprotein coding genes, reflecting significant enrichment; 3) Paternal eQTL are enriched with GWAS hits, and genes harboring paternal eQTL are enriched with drug targets. Our study demonstrates the abundance of POE in gene expression, illustrates the complexity of gene expression regulation, and provides a resource that is complementary to existing resources such as GTEx. We revisited two previous POE findings in light of our POE results. A SNP residing in KCNQ1 that is maternally associated with diabetes is a maternal eQTL of CDKN1C, not KCNQ1. A SNP residing in DLK1 that showed paternal polar overdominance for human obesity is a maternal eQTL of MEG3, offering an explanation for the baseline risk of homozygous samples through association between MEG3 expression and obesity. Finally, we advised caution on conducting Mendelian randomization using gene expression as the exposure.

Introduction

Parent-of-origin effect (POE) is a phenomenon whereby an allele’s effect on a phenotype not only depends on its allelic identity, but also on the parent from whom the allele is inherited (DeChiara et al., 1991; Cockett et al., 1996). POE can be driven by factors such as sex bias in transmission of genetic variants (Tomé et al., 2011) and maternal genetic effects that influences the environment in which the offspring develops (Hager et al., 2008). The main driver, however, is genomic imprinting, which is a phenomenon where one of the two alleles at a locus is functionally silenced by methylation. Importantly, which parental copy of the allele being silenced is highly consistent for that gene across samples. Imprinting appears to play an important role in embryonic and placental development and social behavior (Reik and Walter, 2001; Garfield et al., 2011). Genes underpinning syndromic disorders such as Prader-Willi, Beckwtih-Weidemann, and Angelman show POE (Lawson et al., 2013). The same is also true for common disease phenotypes, such as breast cancer, diabetes, and cardiovascular disease (Kong et al., 2009; Hanson et al., 2013; Mozaffari et al., 2019).

In genetic association studies, particularly genome-wide association studies (GWAS), the absence of POE is usually assumed implicitly. This assumption is largely out of convenience, as in most GWAS, the parent-of-origin of an allele is not directly observable, and not inferable due to lack of first and second degree relatives in the sample. Exceptions include an Icelandic cohort, with their extensively documented pedigrees and the abundance of closely related samples (Kong et al., 2009), a Hutterite pedigree (Mozaffari et al., 2019), the Framingham Heart Study with three-generations of participants (Kannel et al., 1979), and the biobank datasets, for which parent-of-origin can be inferred using distant relatives (Hofmeister et al., 2022, 2025). Another contributing factor to the implicit assumption may be the presumed scarcity of documented POE in humans. Although POE is believed to be associated with a wide range of complex traits and diseases (Lawson et al., 2013), significant associations reported in GWAS setting are of limited quantity. For example, in a Hutterite study, Mozaffari et al. (2019) examined 21 phenotypes and produced 18 significant associations with POE, fewer than one association per phenotype.

Mouse studies demonstrated that, despite a limited number of imprinted genes, most phenotypes display POE, and loci that show POE were not enriched for known imprinted genes. Using reciprocal F1 crosses, Mott et al. (2014) showed that non-imprinted genes can generate POE by interactions with imprinted loci. Focusing on metabolic traits and using a mouse population at different levels of intercrossing, Macias-Velasco et al. (2022) identified a network comprised of three imprinted and six non-imprinted genes that show POE. A recent stduy on pigs demonstrated that a large number genes show POE effect in association with back fat and longissimus dorsi (Li et al., 2025). In humans, however, loci with POE in relation to phenotypes are either near imprinted regions by design (Kong et al., 2009; Hanson et al., 2013), or near regions with characteristic of imprinting from a genome-wide scan (Mozaffari et al., 2019). To what extend POE contributes to phenotypic variation in humans and how widespread the loci with POE in the human genome remains elusive.

Using whole genome sequencing and RNA sequencing data from the Framingham Heart Study, and taking advantage of its relative abundance of trios, we set to investigate to what extent and degree gene expressions is affected by parent-of-origin. There exist multiple studies on POE on gene expression (Zhabotynsky et al., 2019; Jadhav et al., 2019; Deng et al., 2020) or POE of methylation and gene expression jointly (Zink et al., 2018), but these studies are either underpowered due to small sizes or focus on analysis of allele specific gene expression instead of eQTL. A systematic study of POE eQTL is lacking and our study aims to fill this void.

Results

Overview of data processing

We identified 1477 trios with whole genome sequencing (WGS) data and children within the trios having RNAseq data. The trios were selected from participants in the Framing-ham Heart Study (FHS) who are predominantly of European descent. After routine QC of genotype data, for each trio, we first phased the child based on rules of Mendelian inheritance, which resolved all non triple heterozygous genotypes. (See Methods for details on Mendelian errors rate and how to handle them.) We then marked the remaining triple heterozygous genotypes as missing, which produced phased paternal and maternal haplotypes for each child in the trio. Next we imputed missing alleles into their paternal and maternal haplotype background, and the parental haplotype with a higher imputed allele dosage was assigned the reference allele, the other haplotype was assigned the alternative allele. Simulation studies showed this mask and impute phasing to be highly accurate (Supplementary Table S1). Thus at each SNP, we have a genotype vector, a paternal vector, and a maternal vector.

To process RNAseq data, we started from Transcript Per Million (TPM) values, and selected 16, 824 genes that have < 5% missing values (i.e., 0 TMP). We then corrected for the GC content bias using an approach that is similar to the one used by EDAseq (Risso et al., 2011). The corrected gene expression values were then quantile normalized.

Bayes factors as evidence for association

For genetic association we used a linear mixed model, fitted by a novel and efficient method that is designed for analyzing multiomics datasets (Guan and Levy, 2024a). The genetic relatedness matrix (twice of kinship matrix) used for the linear mixed model was estimated by Kindred (Guan and Levy, 2024c). The covariates included age, sex, body mass index, and white blood cell composition. This study focuses on genetic association between gene expression and genetic variants in cis, defined as within 1Mb from transcription start site.

We elected to use Bayes factors (Stephens and Balding, 2009) instead of p-values as the evidence of association for the following considerations. First, at many SNPs, paternal and maternal alleles have different allele frequencies, presumably due to sampling variation. P-values are not comparable unless they have the same power, while the same power requires equal allele frequencies. On the other hand Bayes factors are comparable because they take into account power through prior specifications. Second, Bayes factors panelize genetic variants with small allele frequencies to reduce false positives (Guan and Stephens, 2011; Zhou and Guan, 2018). This is useful as we observed extremely low paternal or maternal allele frequencies. Third, Bayes factors are also convenient in comparing different alternative models, such as paternal effects versus genotype effect, a setting that is difficult for p-values.

For each gene and its cis-SNP, we computed four Bayes factors: genotype Bayes factor (BFg), paternal Bayes factor (BF1), maternal Bayes factor (BF0), and joint Bayes factors (BFj). When computing joint Bayes factor, we also computed paternal and maternal effect sizes and attached a p-value to test whether they differ significantly. Examples of test statistics are provided in Table 1. BFg, BF0 and BF1 are one degree of freedom tests, but BFj is a two degree of freedom test. Thus to achieve BFj > BFg, the paternal and maternal effect sizes have to differ significantly to compensate for the extra degree of freedom. Examples in Table 1 are chosen because their paternal and maternal effects differ significantly, and we have BFj ≫ BFg. Another useful observation is that BF1 is correlated with |β1|, and BF0 is correlated with |β0|. But BF1 and BF0 are oblivious to the signs of β1 and β0, and signs are informative and interesting in our context.

We used log10 BF = 4 as the significance threshold. If the prior odds for a cis-eQTL is 1 out of 1000, this threshold gives the posterior probability of association (PPA) of 0.91. If the prior odds for a cis-eQTL is 1 out of 100, this threshold gives the PPA of 0.99. 1 − PPA is the Bayesian counterpart of the local false discovery rate (c.f. Soloff et al., 2024). We included more discussions regarding p-values, Bayes factors, and FDR in Supplementary.

Overview from sentinel eQTL of joint analysis

Table 1 contains 10 genes and test statistics of their sentinel eQTL. A sentinel eQTL for a gene is defined as the eQTL with largest BFj for that gene. A gene with significant eQTL is called eGene. These sentinel eQTL are either BF1 ≫ BF0 ≈ 1 such as NDN, or BF0 ≫ BF1 ≈ 1 such as MEG3, or BFj ≫ BFg ≈ 1 such as NECAB3. In Table 1 the sentinel eQTL were ordered according to column P, which measures the significance of differences between paternal and maternal effects. Reassuringly, six out of top ten eQTL with most significant P are from bona fide imprinted genes (colored in blue) according to http://geneimprint.com.

Figure 1 plots paternal effect vs maternal effect for all sentinel eQTL whose log10 BFj > 3. The color of dots corresponding to degree of difference between paternal and maternal effects. The marked sentinel eQTL include known imprinted genes such as maternally expressed ZNF331 and MEG3, and paternally expressed FAM50B, GNAS, SGCE, NDN, SNURF, SNRPN, and PEG10. Imprinted genes are located along the x and y-axes. At least two genes appear to be novel imprinted genes: PPIEL and CCR9. Gene ZNF890P provides an example of paternal and maternal effects being in the same direction, but their sizes differ significantly. Finally, NECAB3 and LSM7 are two examples of paternal and maternal effects that are in opposite directions.

Comparison of paternal effects and maternal effects of sentinel eQTL in joint analysis.

Each point is the most prominent eQTL with log10 BFj > 3 of gene. The paternal effect is on x-axis and maternal effect y-axis. The coloring reflects significance of differentials between paternal effects and maternal effects, with red more significant than blue.

This approach based on the sentinel eQTL suggests that 1) the combined effect of paternal and maternal allele can be across all 360 degrees; 2) when paternal and maternal effects are in opposite directions, both effects sizes are modest compared to when they are in the same direction; and 3) there are quite a few sentinel eQTL whose paternal and maternal effects show opposite directions.

NECAB3 : opposite paternal and maternal effects

The existence of eGenes whose eQTL show opposite paternal and maternal effects is intriguing. We focus on the example of NECAB3 to examine further details. Figure 2 left panel shows that all eQTL of NECAB3 have opposite paternal and maternal effects. Consequently, the joint Bayes factors BFj’s are much larger than their corresponding genotype Bayes factors BFg’s (right bottom). We select as an example SNP rs4911348, which is different from the sentinel eQTL shown in Figure 1, and looked into details of its association with the gene expression of NECAB3 (phenotype). Three panels of boxplots (right top) show that genotype has no association with the phenotype, and both paternal alleles and maternal alleles are associated with the phenotype, but in opposite directions. Opposite paternal and maternal effects were also observed in Hutterite POE study, where ten associations of opposite effects across nine traits were reported (Mozaffari et al., 2019).

An example of gene that whose paternal and maternal eQTL have opposite effect sizes.

On the left is paternal vs maternal effect sizes normalized by their corresponding standard deviation. The effect sizes were estimated by joint analysis. The non-significant eQTL are colored in gray and significant eQTL colored in black. An eQTL marked in blue was chosen to show boxplots with gene expression (phenotype) of NECAB3. The three panels of boxplots are for genotypes, paternal alleles, and maternal alleles. For this SNP rs4911348, the genotypes has no association with phenotype, but both paternal alleles and maternal alleles are associated with the phenotype, and the paternal and maternal effects are in opposite direction. Right panel bottom show Manhattan plot of cis-eQTL of the gene NECAB3. The genotype Bayes factor were colored in gray and joint test Bayes factor were colored in black. SNP rs4911348 was highlighted in blue.

POE eQTL are abundant

Next we looked beyond the sentinel eQTL and examined all significant eQTL. We noted that there are 15, 893 eQTL from 14, 733 SNPs and 1, 824 eGenes that are insignificant for the SNP test, but significant for the paternal, or maternal, or joint test. That is, log10 BFg < 4, but either log10 BF1 > 4, or log10 BF0 > 4, or log10 BFj > 4. In other words, these eQTL cannot be detected without analyzing POE, which implies that POE can contribute to recover missing heritability.

The minor allele frequencies of the POE eQTL are balanced. Supplementary Figure S4 empirically demonstrates that our findings are not disproprtionately driven by low-frequency variants and provide a more complete picture of the genetic architecure underlying these POE signals.

With reference Figure 1, we proposed criteria to define subsets of eQTL. The first subset eQTL locate along the x-axis, mimicking sentinel eQTL of NDN and FAM50B ; the second subset locate along the y-axis, mimicking sentinel eQTL of ZNF331 and MEG3 ; the third subset locate along the secondary diagonal, mimicking sentinel eQTL of NECAB3 and LSM7 ; and the fourth subset follows (blues dots) along the diagonal line. Note Figure 1 only involves BFj, our criteria to define gene sets also takes into account of BF1, BF0, and BFg (details in Methods).

The first set of eQTL have significant paternal effects but insignificant maternal effects. We identified 15, 576 such paternal eQTL (set SP) from 14, 372 SNPs associated with 1, 188 eGenes. The second set of eQTL have significant maternal effects but insignificant paternal effects. We identified 14, 783 such maternal eQTL (set SM) from 13, 293 SNPs associated with 1, 209 eGenes. We refer to these paternal and maternal eQTL as imprinting eQTL. The third set of eQTL have opposite paternal and maternal effects, such that joint Bayes factors are much larger than genotype Bayes factor BFj ≫ BFg. We identified 688 such opposing eQTL (set SO) from 485 SNPs that are associated with 180 eGenes. Imprinting eQTL and opposing eQTL are referred to as POE eQTL.

The fourth set of eQTL require that both paternal and maternal alleles alone are associated with an expression phenotype, and the two effects are in the same direction. For these eQTL BFg > BFj because similar effect sizes favor one degree of freedom test. We identified 884, 119 such genotype eQTL from 577, 701 SNPs and 4940 eGenes. The percent of eGenes with genotype eQTL is on par, but smaller than GTEx study (GTEx Consortium, 2020), presumably because log10 BFg > 4 is a more stringent threshold for eQTL analysis.

Paternal eQTL are enriched with GWAS hits

Using ANNOVAR (Wang et al., 2010), we annotated all four sets of eQTL (Table 2). We made the following observations: 1) The median distance to transcription start site (TSS) for opposing eQTL is 673.2Kb, much larger than other types of eQTL, and majority of opposing eQTL SO are located in introns of nearby genes. 2) The percent of GWAS hits for SNPs in SO is 0.022, significantly lower than that of SNPs in genotype eQTL SG’s 0.054 (test for proportion P = 0.001). This is reasonable as GWAS mainly use genotype test and are agnostic to opposing eQTL. 3) The precent of GWAS hits for SNPs in paternal eQTL SP is higher than that of SNPs in SG (test for proportion P = 9 × 10−11), and the percent of GWAS hits among SNPs in maternal eQTL SM is similar to that of SNPs in SG (test for proportion P = 0.051).

Number of eQTL per SNP (r) measures degree of pleiotropy of a set of SNPs. SNPs in SP and SNPs in SM have similar pleiotropy r ≈ 1.1, SNPs in SG have the largest r = 1.53, and SNPs in SO display r = 1.41. In other words, SNPs are more exclusive for imprinting eQTL, less exclusive for genotype eQTL, and somewhere in between for opposing eQTL. Table 2 (bottom) compares how the degree of pleiotropy affects their annotation for SNPs in SG, where set Pk contains SNPs in SG that are eQTL of at least k eGenes. The most interesting observation is that the percent of GWAS hits increases with rising SNP pleiotropy. This feature is only observed in genotype eQTL.

POE eGenes are abundant

There are 1, 188 eGenes harboring paternal eQTL and 1, 209 eGenes harboring maternal eQTL. Their union contains 2, 139 eGenes harbor imprinting eQTL, and their intersection contains 258 eGenes harboring both paternal and maternal eQTL. There are 4, 940 eGenes harboring genotype eQTL, and 1, 867 eGenes harboring both genotype eQTL and imprinting eQTL. Supplementary Figure S1 showed an example of an eGene harboring both maternal and paternal eQTL: GZMH, and an example of an eGene harboring both genotype eQTL and imprinting eQTL: DSE. The most prominent feature is that paternal eQTL and maternal eQTL cluster in separate genomic regions, and genotype eQTL and imprinting eQTL also cluster in separate genomic region.

eGenes that harbor exclusively imprinting eQTL

There are 129 eGenes (set G1) that harbor exclusively paternal eQTL, and 139 eGenes (set G0) that harbor exclusively maternal eQTL. Naturally, G1 contain many bona fide imprinted genes with paternal expression such as FAM50B, NDN, SGCE, SNRPN, SNURF and PEG10. Remaining genes in G1 are candidates for imprinted genes with paternal expression, for example CCR9 and NCOA2. G0 contain bona fide imprinted genes with maternal expression such as GRB10, ZNF331 and MEG3. The remaining in G0 are candidates for imprinted genes with maternal expression, for example COA8 and ZNF888.

The Supplementary Material provides a full list of the candidate imprinted genes from the eQTL analysis. Note however, that not all imprinted genes have eQTL, therefore this list is incomplete. Moreover, a gene on the list is considered behaving like an imprinted gene, but may not be a bona fide imprinted gene. For example, a gene that has no cis-eQTL on its own, may “acquire” a cis-eQTL through gene-gene interaction with a putative imprinted gene such that the cis-eQTL is in fact its trans-eQTL.

We used G1 and G0 as proxies for paternal and maternal imprinted genes, and used cytoBands these genes occupies as proxies for imprinted regions to examine to what extend paternal and maternal eQTL locate in the same cytoBands as these proxy imprinted genes. There are 811 autosome cytoBands in humans, 765 of them larger than 1 Mb, and the median size is 3.2 Mb (Cheung et al., 2001). The 129 proxy paternal imprinted genes occupy 89 cytoBands (set CPI), while 15, 576 paternal eQTL occupy 445 cytoBands (set CPe), among them 9, 947 (or 64%) were not in the cytoBands set CPI. On the other hand, the 139 proxy maternal imprinted genes occupy 96 cytoBands (set CMI), and 14, 783 maternal eQTL occupy 465 cytoBands (set CMe), among them 7, 987 (or 54%) were not in the cytoBand set CMI.

Therefore, a majority of imprinting eQTL are not in the vicinity of the proxy imprinted regions. Interestingly, CPI and CMI share only 14 cytoBands, 8% of their union of 171 cytoBands. As a comparison, CPe and CMe share 349 cytoBands, 62% of their union of 561 cytoBands.

eGenes that harbor exclusively opposing eQTL

There are 99 eGenes (set G2) that harbor only opposing eQTL (Supplementary). Among them 58 encode phosphoproteins (Supplementary), a significant enrichment (FDR = 5.5 × 10−4) according to DAVID, a web server for functional enrichment analysis (Sherman et al., 2022). A phosphoprotein is a protein that is posttranslationally modified by the attachment of either a single phosphate group, or a complex molecule such as 5’-phospho-DNA, through a phosphate group. Because phosphorylation often serves as an on/off switch, targeting phosphoproteins or the enzymes regulating them can restore normal cell signaling in diseases. In addition, drugs can modulate the functions of these proteins, either inhibiting or enhancing their activities. This makes phosphoprotein attractive drug targets. Indeed, according to a therapeutic target database (Zhou et al., 2024), among 58 phosphoproteins genes that harbor exclusively opposing eQTL, there are three successful drug targets ITPR1, ITPR2, and REL, and three additional clinical trial targets CD46, NOTCH3, and XPO1.

Other notable phosphoproteins genes include CDKN2A, PHIP, and LEP, among others. Where CDKN2A produces two major proteins p16(INK4), which is a cyclin-depnednet kinase inhibitor, and p14(ARF), which binds the p53-stabilizing protein MDM2 (Serrano et al., 2000); PHIP is associated with Chung-Jansen syndrome, featuring behavioral problems, intellectual disability, obesity, and dysmorphia (Webster et al., 2016; Jansen et al., 2018); and LEP encodes leptin, a protein that plays a critical role in the regulation of body weight. Leptin is secreted by white adipocytes, and it binds to the leptin receptor in the brain, which in turn inhibits appetite and promotes energy expenditure (Maffei et al., 1995; Weigle et al., 1997).

Paternal eGenes are enriched with drug targets

We used a therapeutic target database (Zhou et al., 2024) to examine the enrichment with drug targets of the four sets of eGenes, grouped by their eQTL. Table 3 contains the enrichment of the four gene sets in either successful targets, clinical trial targets, or combined targets. The significant enrichment that (following Bonferroni correction of 12 tests) is highlighted. Both paternal eGenes and genotype eGenes are enriched in clinical trial targets and combined targets, but paternal eGenes are significantly more enriched than genotype eGenes (one-sided test for proportion P = 0.035). The enrichment of drug targets for paternal eGenes echoes the enrichment of GWAS hits for paternal eQTL (Table 2). This asymmetry between paternal and maternal eGenes perhaps finds its root in the conflicting interests of two sets of genes in relation to transfer of nutrients from the mother to her offspring (Moore and Haig, 1991; Hitchcock and Gardner, 2019). In maze, gene expression of the hybrid is regulated exclusively by the paternally transmitted alleles (Swanson-Wagner et al., 2009).

Discussion

We performed parent-of-origin analysis of eQTL in whole blood gene expression using 1477 trios from the Framingham Heart Study. Some highlights of our study include: 1) There are more than 2,000 genes harboring POE eQTL and majority POE eQTL are not in the vicinity of known imprinted genes; 2) Among 180 genes harboring opposing eQTL, 99 harbor exclusively opposing eQTL, and 58 of the 99 are phosphoprotein coding genes, reflecting significant enrichment; 3) Paternal eQTL are enriched with GWAS hits, and genes harboring paternal eQTL are enriched with drug targets.

Although our study lacks replications in independent cohorts, we have taken several steps to ensure the internal validity and robustness of our findings within the current study. Specifically, we used well-established, bona fide imprinted genes (e.g., MEG3, NDN, SNURF, as listed in Table 1 and Figure 1) as positive control. We used conservative calling creiteria, and explored different parameters (Supplementary Table S2) and chose the one that minimized the inclusion of false positives, ensuring that our core gene sets (e.g., G1, G0) are high-confidence discoveries. We developed a rigorous analytical pipeline, including traio-based phasing validated by simulation (Supplmentary Table S1), the use of linear mixed models to control for relatedness and population structure, and the application of Bayes factors which inheritenly panalize variants with low minor allele frequencies, thereby reducing spurious associations. We believe these internal consistency checks and methodological rigor provide strong confidence in our findings.

Our study demonstrated abundance of POE in gene expression, illustrated additional complexity of gene expression regulation, and provided a resource that is complementary to existing resources such as GTEx. We conclude by revisiting two previous findings in light of our work, and advising caution on conducting Mendelian randomization using gene expression as the exposure.

T2D and maternal allele of rs2237892

SNP rs2237892 in the last intron of KCNQ1 was found in association with type 2 diabetes (T2D) in a Japanese cohort (Yasuda et al., 2008), and the association was later confirmed in a Chinese cohort (Liu et al., 2009). Three other SNPs (rs2283228, rs2237895, and rs2237897) in the same last intron of KCNQ1 were found in association with T2D in another Japanese cohort (Unoki et al., 2008). Both rs2237895 and rs2237897 were replicated in a Singaporean cohort, a Danish cohort (Unoki et al., 2008), and a Chinese cohort (Liu et al., 2009). In these studies, KCNQ1 is identified as T2D candidate gene.

Based on these results from the genotype test, Kong et al. (2009) demonstrate that SNP rs2237892 is maternally associated with T2D in an Iceland cohort. This maternal association was replicated with a larger odds ratio in a Pima Amerindian cohort (Hanson et al., 2013). According to data from GTEx, none of these SNPs are eQTL of KCNQ1 and CDKN1C in any tissue, consistent with the fact that these are imprinted genes and GTEx study is oblivious to parent of origin. Our eQTL analysis showed that SNPs rs2237892, rs2283228, and rs2237897 are maternal eQTL of CDKN1C, a down-stream neighbor of KCNQ1, and none of these SNPs are eQTL of KCNQ1.

We therefore suggest that CDKN1C, instead of KCNQ1, is a T2D candidate gene, for the following reasons: 1) These GWAS SNPs connect T2D and CDKN1C quantitatively, but connect T2D and KCNQ1 only geographically; 2) A study suggests that CDKN1C mutations may represent a novel monogenic form of diabetes (Kerns et al., 2014); 3) A boy carrying a frameshift mutation in CDKN1C was diabetic from week 29 (Berland et al., 2022); and 4) Targeted demethylation at the CDKN1C/p57 locus induces human β cell replication, while the loss of insulin-secreting β cell is characteristic among T1D and T2D Ou et al. (2019).

We note that gene expression in our study is from whole blood. While whole blood is a valuable and accessible tissue, replication in T2D-relevant tissues (e.g., pancreas, adipose) would be an important future direction, and our findings provide a hypothesis for such targeted investigations.

Polar overdominance and rs1802710

Heterozygous overdominance is a pattern of inheritance such that both homozygous AA and BB have the same baseline trait value, while heterozygous AB has a higher trait value. Polar overdominance (Cockett et al., 1996) introduces asymmetry into overdominance to separates AB into two groups according to parent-of-origin of A (or B), such that AB with paternal A (or B) has a high trait value, while AB with maternal A (or B) has the baseline trait value.

Wermter et al. (2008) studied trios of extremely obese offspring and identified rs1802710 in exon 5 of DLK1, homologous to the ovine callipyge locus (Cockett et al., 1996), whose allelic transmission pattern was consistent with polar overdominance: frequent transmission of the paternal C allele to obese children, but the relative risk for carriers of the homozygous CC genotype was not increased compared to the reference TT genotype.

Our eQTL results show that DLK1 has no significant eQTL, but the maternal C allele of rs1802710 reduces expression of MEG3, 53Kb downstream of DLK1. Both MEG3 and DLK1 are imprinted with MEG3 maternally expressed and DLK1 paternally expressed. The expression of MEG3 was shown to be significantly higher in the obese group (Danesh-moghadam et al., 2021). Therefore the maternal C allele is associated with reduced risks of obesity, which balances out the increased risk conferred by paternal C allele, thus offering an explanation for the baseline risk of homozygous CC samples in (Wermter et al., 2008). Since rs1802710 is not an eQTL of any gene in any tissue according to GTEx data, our study is critical to link rs1802710 with MEG3, albeit not in the adipose tissue.

Mendelian randomization

Mendelian randomization (MR) refers to the random allocation of alleles at the time of gamete formation. Observational epidemiology studies use MR to infer the causal effect of an exposure on an phenotype (Smith and Ebrahim, 2003), as if the random allocation of alleles is comparable to a randomized clinical trial. An important assumption for MR, among many other important assumptions, is that the phenotype conferred by a specific genetic variant is homogeneous in the population, and exchangeable between paternally and maternally inherited alleles. POE is a direct violation of this assumption (Bochud et al., 2008).

There is a growing interest in using gene expression as an exposure to perform MR across various traits (Porcu et al., 2019; van der Graaf et al., 2020). If an exposure has eQTL with POE, ignoring the parent-of-origin will bias prediction of a subset of samples. If the bias is severe it will change the ranking of the exposure. Consequently, it either compromises the power of the MR or leads to a false conclusion of a causal relationship between the exposure and the phenotype.

Our study demonstrates the abundance of POE in gene expressions, and we therefore advise caution when conducting MR using gene expression as the exposure. We suggest checking the list of POE SNPs and eGenes we provided in the Supplementary Material and exclude those that show POE towards the exposure. The same suggestion is also applicable to computing polygenic risk scores, and imputing gene expression such as PrediXcan (Gamazon et al., 2015) and a Bayesian method motivated by it (Qi et al., 2018).

Method

Genotype data

We first used pedigree and availability of RNAseq data to narrow the samples down to 2955, and performed routine QC to obtain bi-allelic SNPs (Taliun et al., 2021). We then used Kindred to infer pairwise kinship for each chromosome, and computed mean and sd of the kinship. A pair of sample that is parent-offspring can be distinguished from full sibs by the sd (Supplementary Figure S2). Trios are consisted of sample A, B, and C such that ϕ(A, B) ≈ 0 and ϕ(A, C) ≈ ϕ(B, C) ≈ 0.25 where ϕ(·, ·) is the kinship. In the end, we identified 1477 trios that have whole genome sequencing data and whose children have RNAseq data. Totally, we have 7, 752, 281 autosome bi-allelic SNPs with minor allele frequency > 0.01 (not corrected for relatedness). Across all 1477 trios, the Mendelian error is negligible 4.76 per 100, 000 SNPs per trio. To handle Mendelian errors, we assume child was correctly genotyped, and errors happened in parents, but if the child is heterozygous, we assume this is a triple heterozygous SNP. On average 1477 trios have 65 triple heterozygous genotypes per SNP. The triple heterozygous SNPs cannot be phased by Mendelian inheritance, but can be phased based on a linkage disequilibrium model (below).

Inference of parent of origin

For each trio, we first phase all markers that are not triple heterozygous using rule of Mendelian inheritance, where for the child in each trio we obtain paternal and maternal haplotypes with triple heterozygous markers whose phase are unresolved. We mark those triple heterozygous markers as missing. For markers that show Mendelian incompatibility, we assume child’s genotype is correct and mark as missing if it is heterozygous. Using a hidden Markov model that was designed to model haplotype variation (Guan, 2014), we impute missing markers in each haplotypes to obtain dosage estimates (between 0 and 1). Then for each marker that are marked as missing, we compare imputed dosages between paternal and maternal haplotypes, and assign the large dosage as 1 and small dosage as 0. We thus obtain paternal and maternal haplotypes with no missing data. We simulated trios using haplotypes from 1000 Genomes project, and investigated the accuracy of the phasing by mask and imputation approach described above. The phasing is highly accurate, the error rate is less than 1 out of 200 triple heterozygous SNPs (Supplementary Table S1).

RNAseq data

The RNAseq data of 1477 chilren comes from three batches with 1381 comes from batch 1; 14 comes from batch 2, and 82 comes from batch 3. Our primary goal is to correct for GC content bias. To this end, we first remove genes whose proportion of 0 TPM values is greater than 5%, which retain 16, 969 genes out of total 58, 103 transcripts, majority of which are coding genes. Among those, 16, 824 genes have GC content. To correct for GC content bias, we took an approach used by EDAseq to regress out the GC content from log(1+TMP), but instead of using local linear regression as documented in EDAseq, we used local quadratic regression implemented in loess in R. Specifically, let y = log(1+TMP) be a gene expression and g be the GC content, we fit ly = loess(y, g) and compute ŷ = median(y) + ly$residuals. We then quantile normalize ŷ by qqnorm(ŷ, plot.it = F)$y separately for each gene.

eQTL analysis

We used IDUL to perform eQTL analysis. IDUL fits linear mixed models to achieve exact optimal. It was specifically developed to analyze multi-omics data to achieve high efficiency by reusing the intermediate computations (Guan and Levy, 2024a). Linear mixed model includes a random effect with a covariance structure defined by the genetic relatedness matrix (GRM). This approach is widely regarded as more robust than including a limited number of principal components, as it accounts for both fine-scale population stratification and known relatedness simultaneously. Previous analysis showed that for Framingham heart study, the genomic inflation was well controlled using linear mixed model with kinship matrix computed by Kindred (Guan and Levy, 2024c). In this study, we also controlled age, sex, BMI, and cell compositions in peripheral blood. We now briefly describe how Bayes factors were computed.

Consider a model (1) y=Wa+xβ+Zu+eu∼MVNn(0,τ−1ηK)e∼MVNn(0,τ−1In) where W contains conventional covariates such as age and sex, including a column of 1, x contains genetic variant(s) to be tested for association, u is the random effect with Z as its loading matrix and twice of kinship K as its covariance (both Z and K are known), MVNn denotes an n-dimensional multivariate normal distribution, In is n-dimensional identity matrix. In our anlaysis Z is the identity matrix. Denote X = (W, x) and b = (a, β), then Xb is the fixed effect, and we assume X has a full rank c. In genetic association studies, the random effect Zu is a nuisance term that absorbs part of the phenotype y that is attributable to population stratification and relatedness. The maximum likelihood estimate (MLE) of η can be efficiently obtained (Guan and Levy, 2024a), plug η^ back into (1), and specify the following conjugate prior (2) a∼N(0,τ−1Va)β∼N(0,τ−1σ2)τ∼Γ(κ1/2,κ2/2) and let Va → ∞, κ1 → 0 and κ2 → 0, Bayes factor can be evaluated efficiently in a closed form. This study we used σ = 0.5, following prior work of Bayes factors for linear models (Servin and Stephens, 2007). Complete details on computing Bayes factors for linear mixed model can be found in (Guan and Levy, 2024b).

P-value for difference between paternal maternal effects

For joint analysis, we have (3) y=Wa+x1β1+x2β2+Zu+e with the prior for β1 and β2 as (4) β1∼N(0,τ−1σ2)β2∼N(0,τ−1σ2) The Bayes factor can be similarly computed in a closed form. We obtain the posterior estimates of β^1 and β^2, their variances s12 and s22 and covariance s12. To test null hypothesis β^1=β^2, we compute test statistics t=(β^1−β^2)2s12+s22−2s12. Since under the null t following χ2 distribution with 1 degree of freedom, we can compute a p-value.

Threshold for eQTL sets and gene sets

For paternal eQTL set SP, we require log10 BF1 > 4 and log10 BF0 < θ; For maternal eQTL set SM, we require log10 BF0 > 4 and log10 BF1 < θ; For opposing eQTL set SO, we require log10 BFj − log10 BFg > 4 and log10 BF1 > θ and log10 BF0 > θ, and β1 ∗ β0 < 0; For genotype eQTL set SG, we require log10 BFg > 4 and log10 BF1 > θ and log10 BF0 > θ, and β1 ∗ β0 > 0. It’s easy to see for a particular type of eQTL, different θ produce nested eQTL sets. We tried θ = 0, log10 2, and log10 3 to obtain eQTL sets (Supplementary Table S2), from which we obtain gene sets harboring those eQTL GP, GM, GA and GG, from which we obtain genes harbor exclusively paternal eQTL G1 = GP \ (GM ∪ GA ∪ GG), genes harbor exclusive maternal eQTL G0 = GM \ (GP ∪ GA ∪ GG), and genes harbor exclusively opposing eQTL G0 = GA \ (GP ∪ GM ∪ GG). We compare known imprinted genes with paternal expression with G1, and known imprinted genes with maternal expression with G0. θ = log10 2 produced minimum G1 and G0 that contain known imprinted genes.